% !TEX root = ./main.tex

We describe two implementations of the Filtered Stepper Calculus: a
compact reference implementation in OCaml that directly mirrors the
formal rules, and the full integration into the Hazel live programming
environment.

\subsection{Reference Implementation}\label{sec:impl-mini}

To validate the formal semantics presented in Section~\ref{sec:filter},
we built a standalone reference stepper in OCaml. It implements a
substitution-based small-step evaluator for the simply typed lambda
calculus extended with a fixpoint operator and the full filter calculus.
The core is roughly 560 lines of OCaml, structured as a direct
transliteration of the formal judgments: expressions, patterns, actions,
and gas are mutually recursive algebraic data types; filter and residue
expressions are constructors of the expression type.

A single stepping function orchestrates the pipeline described in
Section~\ref{sec:filter_example}, calling instrumentation, optimization,
decomposition, annotation, and decay in sequence. If the annotation for
the selected redex is \(\ActSkip\), the stepper performs the instruction
transition and recurses; if it is \(\ActStep\), it returns the list of
available redexes to the caller.

Because the reference implementation uses substitution-based evaluation
and named variables, it corresponds closely to the on-paper calculus and
is easy to audit against the formal rules. We test the stepper against
known results (e.g., verifying that \texttt{fac(5)} evaluates to 120)
and use it as a test oracle for the Hazel integration: any divergence
between the two indicates a bug in the production stepper.

\subsection{Hazel Integration}\label{sec:impl-hazel}

Hazel internally uses a big-step, environment-based evaluator for
performance: it avoids redundant substitutions and enables tail-call
optimization. However, because the course teaches substitution-based
equational semantics, the stepper must present traces as if
substitution were performed eagerly.

We achieve this by maintaining a single set of semantic transition
rules that both the big-step evaluator and the small-step stepper
interpret. The stepper works over Hazel's internal expression
representation, which includes closures (pairs of an expression and
an environment). During decomposition, closures are traversed like
ordinary expressions, with the environment carried in the evaluation
context. When the stepper displays a step to the user, a
post-processing pass replaces all bound variables with their
environment bindings, producing output that looks substitution-based.

The filter system in Hazel mirrors the reference implementation but
operates over Hazel's richer expression language, which includes
pattern matching, sum and product types, type ascriptions, and typed
holes. Filter matching uses a semantic equality check with special
accommodations: fixpoint wrappers and type ascriptions are ignored
during comparison, and empty holes in a pattern act as the
\(\PatValue\) wildcard.

\subsection{Performance}\label{sec:impl-perf}

In Hazel, the filter-instrumented stepper is more expensive than the
plain big-step evaluator, because every step requires a full
instrumentation pass over the expression tree followed by
decomposition. We applied several techniques to keep interactive
stepping responsive in the Hazel editor.

\paragraph{Optimization of residue nodes.}
After instrumentation, immediately adjacent residue nodes are common
(e.g., when an outer \(\PatExpr\) filter and a more specific inner
filter both match the same sub-expression). An optimization pass
collapses nested residues by retaining only the highest-priority one,
reducing the depth of the expression tree before decomposition.

\paragraph{Physical equality.}
During filter matching, we first check physical equality between the
pattern and the expression before falling back to structural
comparison. Since instrumentation is applied repeatedly across steps,
many sub-expressions are shared and this check short-circuits a large
fraction of comparisons.

\paragraph{Locally nameless representation.}
To test alpha-equivalence of lambda bodies during matching (the
M-Lam and M-Fix rules), we convert expressions to a locally nameless
(de Bruijn) representation. This avoids the cost of repeated
alpha-renaming and provides a canonical form for comparison.

\paragraph{Settings-based pre-filtering.}
In addition to the user-written filter language, Hazel provides UI
toggles that control the visibility of common step kinds: variable
lookups, fixpoint unrolling, type ascription elimination, and case
clause selection. When a step kind is toggled off, the stepper skips
it before filter matching is attempted, avoiding unnecessary work.

\paragraph{Web worker isolation.}
The stepper runs in a dedicated Web Worker thread, so that
long-running step computations do not block the editor's main thread.
The worker communicates with the UI via message passing, and the user
can cancel a stepping computation at any time.

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: "main"
%%% End:
